\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-08-26 paper.qmd
Updated the section 7.2

In the field of single-cell studies, a common analytical task involves clustering to identify groups of cells with similar expression profiles. NLDR is commonly used to display clusters, and help to verify the results. For example, @chen2023 illustrates the use of UMAP to identify clusters in Human Peripheral Blood Mononuclear Cells (PBMC3k). @fig-umap-author is a reproduction of the published plot. It shows three well-separated clusters with big separations.

@@

To determine whether the UMAP representation with the (hyper-)parameter choice suggested by @chen2023 preserves the original data structure, we visualize the model constructed with UMAP overlaid on the $9\text{-}D$ data. The @fig-umap-author shows three well-separated clusters. However, as shown in @fig-pbmc1-sc, there is no big separation between three clusters in $9\text{-}D$. Therefore, the suggested UMAP representation (@fig-umap-author) does not accurately represent the structure(s) present in PBMC3k dataset. But, when visualizing the model-in-th-data-space some unobserved structures can be seen. Some clusters have non-linear continuity patterns and high-density patches (@fig-pbmc1-sc).

@@

PaCMAP provides a reasonable \gD{} representation of MNIST digit 1 data, as it preserves the non-linear structure present in the $10\text{-}D$ data (@fig-mnist-tri-proj). Furthermore, shows some quirks when visualizing the *model-in-the-data-space* (@fig-mnist-tri-proj).

@@

```{r}
#| label: fig-model-mnist
#| fig-cap: "$10\\text{-}D$ model error in PaCMAP layout of MNIST digit 1 dataset in $2\\text{-}D$. Most low $10\\text{-}D$ model errors are distributed along the lower edge of the $2\\text{-}D$ structure, while most large $10\\text{-}D$ model errors are concentrated along the upper edge (a). There are certain data points that exhibit large error rates due to their deviation from the usual $10\\text{-}D$ data structure, which makes them anomalies. These anomalies can be classified into two types: those that are anomalies within the non-linear structure and those that lie outside of it. The images associated with high model error points within the non-linear structure display different patterns of the digit 1 (b). However, when comparing these images to the ones found outside of the non-linear structure, it becomes evident that the latter display different patterns of the digit 1 (c). "
#| fig-pos: H
#| fig-width: 4
#| fig-height: 8
#| out-height: 80%

>>>>

In the field of single-cell studies, a common analytical task involves clustering to identify groups of cells with similar expression profiles. NLDR methods are commonly used to display clusters, and help to verify the results. For example, @chen2023 illustrates the use of UMAP to identify clusters in Human Peripheral Blood Mononuclear Cells (PBMC3k). @fig-umap-author is the reproduction of the published plot. It shows three well-separated clusters with big separations.

@@

To determine whether the UMAP representation with the specific (hyper-)parameter choice suggested by @chen2023 preserves the original data structure, we visualize the model constructed with UMAP overlaid on the $9\text{-}D$ data. The @fig-umap-author shows three well-separated clusters. However, as shown in @fig-pbmc1-sc, there is no big separation between three clusters in $9\text{-}D$. Therefore, the suggested UMAP representation (@fig-umap-author) does not accurately represent the structure(s) present in PBMC3k dataset. But, when visualizing the model-in-th-data-space some unobserved structures can be seen. Some clusters have non-linear continuity patterns and high-density patches (@fig-pbmc1-sc).

@@

As a result of visualizing the *model-in-the-data-space*, PaCMAP provides a reasonable \gD{} representation of MNIST digit 1 data, as it preserves the non-linear structure present in the $10\text{-}D$ data (@fig-mnist-tri-proj). It also shows some quirks like twisted patterns and long edges (@fig-mnist-tri-proj). Furthermore, the results help to identify anomalies (with large model errors) within and the outside the non-linear structure, displaying different patterns of the digit 1 (@fig-model-error-mnist).

@@

```{r}
#| label: fig-model-error-mnist
#| fig-cap: "$10\\text{-}D$ model error in PaCMAP layout of MNIST digit 1 dataset in $2\\text{-}D$. Most low $10\\text{-}D$ model errors are distributed along the lower edge of the $2\\text{-}D$ structure, while most large $10\\text{-}D$ model errors are concentrated along the upper edge (a). There are certain data points that exhibit large error rates due to their deviation from the usual $10\\text{-}D$ data structure, which makes them anomalies. These anomalies can be classified into two types: those that are anomalies within the non-linear structure and those that lie outside of it. The images associated with high model error points within the non-linear structure display different patterns of the digit 1 (b). However, when comparing these images to the ones found outside of the non-linear structure, it becomes evident that the latter display different patterns of the digit 1 (c). "
#| fig-pos: H
#| fig-width: 4
#| fig-height: 8
#| out-height: 80%

### 2024-08-26 paper.tex
Updated the section 7.2

In the field of single-cell studies, a common analytical task involves
clustering to identify groups of cells with similar expression profiles.
NLDR is commonly used to display clusters, and help to verify the
results. For example, \citet{chen2023} illustrates the use of UMAP to
identify clusters in Human Peripheral Blood Mononuclear Cells (PBMC3k).
\textbf{?@fig-umap-author} is a reproduction of the published plot. It
shows three well-separated clusters with big separations.

To determine whether the UMAP representation with the (hyper-)parameter
choice suggested by \citet{chen2023} preserves the original data
structure, we visualize the model constructed with UMAP overlaid on the
\(9\text{-}D\) data. The \textbf{?@fig-umap-author} shows three
well-separated clusters. However, as shown in \textbf{?@fig-pbmc1-sc},
there is no big separation between three clusters in \(9\text{-}D\).
Therefore, the suggested UMAP representation
(\textbf{?@fig-umap-author}) does not accurately represent the
structure(s) present in PBMC3k dataset. But, when visualizing the
model-in-th-data-space some unobserved structures can be seen. Some
clusters have non-linear continuity patterns and high-density patches
(\textbf{?@fig-pbmc1-sc}).

@@

PaCMAP provides a reasonable \gD{} representation of MNIST digit 1 data,
as it preserves the non-linear structure present in the \(10\text{-}D\)
data (Figure~\ref{fig-mnist-tri-proj}). Furthermore, shows some quirks
when visualizing the \emph{model-in-the-data-space}
(Figure~\ref{fig-mnist-tri-proj}).

@@

\includegraphics[width=1\textwidth,height=0.8\textheight]{paper_files/figure-pdf/fig-model-mnist-1.pdf}

@@

\caption{\label{fig-model-mnist}\(10\text{-}D\) model error in PaCMAP
layout of MNIST digit 1 dataset in \(2\text{-}D\). Most low
\(10\text{-}D\) model errors are distributed along the lower edge of the
\(2\text{-}D\) structure, while most large \(10\text{-}D\) model errors
are concentrated along the upper edge (a). There are certain data points
that exhibit large error rates due to their deviation from the usual
\(10\text{-}D\) data structure, which makes them anomalies. These
anomalies can be classified into two types: those that are anomalies
within the non-linear structure and those that lie outside of it. The
images associated with high model error points within the non-linear
structure display different patterns of the digit 1 (b). However, when
comparing these images to the ones found outside of the non-linear
structure, it becomes evident that the latter display different patterns
of the digit 1 (c).}

>>>>

In the field of single-cell studies, a common analytical task involves
clustering to identify groups of cells with similar expression profiles.
NLDR methods are commonly used to display clusters, and help to verify
the results. For example, \citet{chen2023} illustrates the use of UMAP
to identify clusters in Human Peripheral Blood Mononuclear Cells
(PBMC3k). \textbf{?@fig-umap-author} is the reproduction of the
published plot. It shows three well-separated clusters with big
separations.

To determine whether the UMAP representation with the specific
(hyper-)parameter choice suggested by \citet{chen2023} preserves the
original data structure, we visualize the model constructed with UMAP
overlaid on the \(9\text{-}D\) data. The \textbf{?@fig-umap-author}
shows three well-separated clusters. However, as shown in
\textbf{?@fig-pbmc1-sc}, there is no big separation between three
clusters in \(9\text{-}D\). Therefore, the suggested UMAP representation
(\textbf{?@fig-umap-author}) does not accurately represent the
structure(s) present in PBMC3k dataset. But, when visualizing the
model-in-th-data-space some unobserved structures can be seen. Some
clusters have non-linear continuity patterns and high-density patches
(\textbf{?@fig-pbmc1-sc}).

@@

As a result of visualizing the \emph{model-in-the-data-space}, PaCMAP
provides a reasonable \gD{} representation of MNIST digit 1 data, as it
preserves the non-linear structure present in the \(10\text{-}D\) data
(Figure~\ref{fig-mnist-tri-proj}). It also shows some quirks like
twisted patterns and long edges (Figure~\ref{fig-mnist-tri-proj}).
Furthermore, the results help to identify anomalies (with large model
errors) within and the outside the non-linear structure, displaying
different patterns of the digit 1 (Figure~\ref{fig-model-error-mnist}).

@@

\includegraphics[width=1\textwidth,height=0.8\textheight]{paper_files/figure-pdf/fig-model-error-mnist-1.pdf}

@@

\caption{\label{fig-model-error-mnist}\(10\text{-}D\) model error in
PaCMAP layout of MNIST digit 1 dataset in \(2\text{-}D\). Most low
\(10\text{-}D\) model errors are distributed along the lower edge of the
\(2\text{-}D\) structure, while most large \(10\text{-}D\) model errors
are concentrated along the upper edge (a). There are certain data points
that exhibit large error rates due to their deviation from the usual
\(10\text{-}D\) data structure, which makes them anomalies. These
anomalies can be classified into two types: those that are anomalies
within the non-linear structure and those that lie outside of it. The
images associated with high model error points within the non-linear
structure display different patterns of the digit 1 (b). However, when
comparing these images to the ones found outside of the non-linear
structure, it becomes evident that the latter display different patterns
of the digit 1 (c).}
